---
title: "Fluidigm Analysis"
author: "Timothy Smyth"
date: "2024-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fluidigm Analysis

### This script imports, analyzes, and visualizes the results of the Fluidigm 48 gene pannel across indicated polarization states and DEP exposures.

Prepare the environment

```{r}
rm(list = ls(all.names = TRUE)) # clears global environ.

# Load packages
library(tidyverse) # for data cleaning
library(pheatmap) # for making heatmap
library(ggfortify) # for PCA
library(factoextra) # for PCA 
library(ggplot2) # for plotting
library(RColorBrewer) # for color palette
library(ggpubr) # for statistics and graphing
library(broom) # for ANOVA
```

### Import data and format

```{r}
# Import data
Fluidigm <- read.csv("Fluidigm_Data_for_R.csv", fileEncoding="UTF-8-BOM")
Fluidigm_df <- data.frame(Fluidigm, row.names = Fluidigm$Full.Name)
Fluidigm_df <- Fluidigm_df[, 3:ncol(Fluidigm_df)]

# Isolate Vehicle samples
Veh_df <- subset(Fluidigm, Group == 'Vehicle') 
Veh_df <- data.frame(Veh_df, row.names = Veh_df$Full.Name)
Veh_df <- Veh_df[, 3:ncol(Veh_df)]

# Isolate DEP samples
DEP_df <- subset(Fluidigm, Group == 'DEP') 
DEP_df <- data.frame(DEP_df, row.names = DEP_df$Full.Name)
DEP_df <- DEP_df[, 3:ncol(DEP_df)]

# Remove periods from column names
names(Fluidigm) <- gsub("\\.", "", names(Fluidigm))
names(Fluidigm_df) <- gsub("\\.", "", names(Fluidigm_df))
names(Veh_df) <- gsub("\\.", "", names(Veh_df))
names(DEP_df) <- gsub("\\.", "", names(DEP_df))
```

### Create principal component analysis graphs for all data, Vehicle data only, and DEP data only.

```{r}
# Isolate target columns
Fluidigm_PCA <- Fluidigm_df[, 5:ncol(Fluidigm_df)]
Veh_PCA <- Veh_df[, 5:ncol(Veh_df)]
DEP_PCA <- DEP_df[, 5:ncol(DEP_df)]

# Run PCA
pca.res <- prcomp(Fluidigm_PCA, center = TRUE, scale = TRUE)
Veh.pca.res <- prcomp(Veh_PCA, center = TRUE, scale = TRUE)
DEP.pca.res <- prcomp(DEP_PCA, center = TRUE, scale = TRUE)

# Set theme
theme_set(theme_bw())

# PCA for all groups
flu_pca <- fviz_pca_ind(pca.res,
             label = "none",
             habillage = Fluidigm_df$Treatment, 
             palette = "Dark2",
             addEllipses = TRUE) + 
  ggtitle("PCA - Clustering") +
  theme(axis.title = element_text(size = 14), 
        axis.text = element_text(size = 12), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        panel.border = element_rect(fill = NA, color = "black", size = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.text = element_text(size = 14)) +
  xlim(-10, 10)

file_name = 'flu_pca.tiff'
tiff(file_name, units="in", width=8, height=8, res=300)
print(flu_pca)
dev.off()

# PCA for Vehicle exposed groups only
fviz_pca_ind(Veh.pca.res,
             label = "none",
             habillage = Veh_df$Treatment, 
             palette = "Dark2",
             addEllipses = TRUE) + 
  ggtitle("PCA - Clustering") +
  theme(axis.title = element_text(size = 14), 
        axis.text = element_text(size = 12), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        panel.border = element_rect(fill = NA, color = "black", size = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.text = element_text(size = 14)) +
  xlim(-10, 10)

# PCA for DEP exposed groups only
fviz_pca_ind(DEP.pca.res,
             label = "none",
             habillage = DEP_df$Treatment, 
             palette = "Dark2",
             addEllipses = TRUE) + 
  ggtitle("PCA - Clustering") +
  theme(axis.title = element_text(size = 14), 
        axis.text = element_text(size = 12), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        panel.border = element_rect(fill = NA, color = "black", size = 0.3),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.text = element_text(size = 14)) +
  xlim(-10, 10)
```

### Create heatmaps of Vehicle or DEP exposed samples

```{r}
# Create summary data frame with means by group
Veh_mean_df = Veh_df %>% group_by(Treatment) %>% summarize(across(everything(), mean))
Veh_mean_df <- data.frame(Veh_mean_df[5:ncol(Veh_mean_df)], row.names = Veh_mean_df$Treatment)
Veh_mean_df <- Veh_mean_df[c(2, 4, 1, 3), ]

DEP_mean_df = DEP_df %>% group_by(Treatment) %>% summarize(across(everything(), mean))
DEP_mean_df <- data.frame(DEP_mean_df[5:ncol(DEP_mean_df)], row.names = DEP_mean_df$Treatment)
DEP_mean_df <- DEP_mean_df[c(2, 4, 1, 3), ]

flu_veh <- pheatmap(t(Veh_mean_df), 
         color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100), # sets color scheme
         display_numbers = TRUE,
         number_color = "black",
         fontsize_number = 8,
         angle_col = c("45"), # makes column labels horizontal
         cellwidth = 75, # sets dimensions of cells so that they don't change with viewer pane size
         cellheight = 10, # sets dimensions of cells so that they don't change with viewer pane size
         border_color = "black", # adds black border around cells
         treeheight_col = 10, # sets dims of trees so that they don't change with viewer pane size
         fontsize_row = 9, # sets dims of trees so that they don't change with viewer pane size
         scale = 'row', # scales data by row
         fontsize_col = 12, # sets font size for column labels
         cutree_rows = 5, # indicates how many clusters to show separated by spaces
         cluster_cols = FALSE, 
         cluster_rows = TRUE)

file_name = 'flu_veh.tiff'
tiff(file_name, units="in", width=8, height=8, res=300)
print(flu_pca)
dev.off()

flu_dep <- pheatmap(t(DEP_mean_df), 
         color = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100), # sets color scheme
         display_numbers = TRUE,
         number_color = "black",
         fontsize_number = 8,
         angle_col = c("45"), # makes column labels horizontal
         cellwidth = 75, # sets dimensions of cells so that they don't change with viewer pane size
         cellheight = 10, # sets dimensions of cells so that they don't change with viewer pane size
         border_color = "black", # adds black border around cells
         treeheight_col = 10, # sets dims of trees so that they don't change with viewer pane size
         fontsize_row = 9, # sets dims of trees so that they don't change with viewer pane size
         scale = 'row', # scales data by row
         fontsize_col = 12, # sets font size for column labels
         cutree_rows = 5, # indicates how many clusters to show separated by spaces
         cluster_cols = FALSE, 
         cluster_rows = TRUE)

file_name = 'flu_dep.tiff'
tiff(file_name, units="in", width=8, height=8, res=300)
print(flu_pca)
dev.off()
```

### Perform 2-way RM ANOVA with fisher's LSD and export results

```{r}
# Create data frame with mean values of each target for testing of differences
mean_df <- Fluidigm_df %>% group_by(Treatment) %>% summarize(across(everything(),mean))
mean_df <- data.frame(mean_df[5:ncol(mean_df)], row.names = mean_df$Treatment)

# Arrange rows as listed
mean_df <- mean_df %>% arrange(factor(rownames(mean_df), 
                                      c("M0_Vehicle", "M2_Vehicle", 
                                        "M0 -> M1", "M2 -> M1",
                                        "M0_DEP", "M2_DEP", 
                                        "M0 -> M1+DEP", "M2 -> M1+DEP")))

# Create row combinations to find differences
Treatment1 <- seq(1,nrow(mean_df)) 
Treatment2 <- Treatment1

# Calculate differences in groups
mean_differences <- expand.grid(Treatment1, Treatment2) %>% # Expand to match sequences
  filter(Var1 > Var2) %>% # Keep rows where Var1 > Var2
  mutate(abs(mean_df[Var1, ] - mean_df[Var2, ])) # Calculate abs value of Var1 - Var2

# Replace var with Treatment name
mean_differences <- mean_differences %>% mutate(Var2 = 
                                                  case_when(str_detect(Var2, '1') ~ "M0_Vehicle",
                                                            str_detect(Var2, '2') ~ "M2_Vehicle",
                                                            str_detect(Var2, '3') ~ "M0 -> M1",
                                                            str_detect(Var2, '4') ~ "M2 -> M1",
                                                            str_detect(Var2, '5') ~ "M0_DEP", 
                                                            str_detect(Var2, '6') ~ "M2_DEP", 
                                                            str_detect(Var2, '7') ~ "M0 -> M1+DEP", 
                                                            str_detect(Var2, '8') ~ "M2 -> M1+DEP"))

mean_differences <- mean_differences %>% mutate(Var1 = 
                                                  case_when(str_detect(Var1, '1') ~ "M0_Vehicle",
                                                            str_detect(Var1, '2') ~ "M2_Vehicle",
                                                            str_detect(Var1, '3') ~ "M0 -> M1",
                                                            str_detect(Var1, '4') ~ "M2 -> M1",
                                                            str_detect(Var1, '5') ~ "M0_DEP", 
                                                            str_detect(Var1, '6') ~ "M2_DEP", 
                                                            str_detect(Var1, '7') ~ "M0 -> M1+DEP", 
                                                            str_detect(Var1, '8') ~ "M2 -> M1+DEP"))

# Isolate comparisons
comparisons <- mean_differences[1:2]
colnames(comparisons) <- c('group1', 'group2')

# Set row names to Treatments compared in row
mean_differences <- data.frame(mean_differences[3:ncol(mean_differences)], 
                               row.names = paste0(mean_differences$Var2, " : ", mean_differences$Var1))

# Create vectors for storing results of stats
Data_Test <- vector('list', ncol(Fluidigm))
anova <- vector('list', ncol(Fluidigm))
MSE <- vector('list', ncol(Fluidigm))
t_ratio <- vector('list', ncol(Fluidigm))
p_val <- vector('list', ncol(Fluidigm))

# Disable scientific notation
options(scipen=999)

# Run Two-Way RM ANOVA with Fisher LSD
# t-ratio of comparison = 

# [Difference between Treatments]/
# sqrt([Mean squares of within group from anova]*(1/n[group1] + 1/n[group2]))

# Individual p-values calculated from the resulting t-ratio with a DF of 8
for(i in 7:ncol(Fluidigm)) {  
  
  # Two-Way RM Anova
  Data_Test[[i]] <- tidy(aov(Fluidigm[, i] ~ Starting * Exposure 
                             + Error(ID / (Starting * Exposure))
                             , data = Fluidigm))
  
  # Isolate anova p-value for graphing
  anova[[i]] <- round(Data_Test[[i]][6, "p.value"], 4)
  
  # Rename column to p
  colnames(anova[[i]])[1] <- 'p'
  
  # Set values rounded to below 0.0001 to <0.0001
  anova[[i]]$p[anova[[i]]$p<0.0001] <- '<0.0001'
  
  # Mean Square Error
  MSE[[i]] <- sqrt(((1/5 + 1/5) * Data_Test[[i]][7,"meansq"]))
  
  # t ratio
  t_ratio[[i]] <- mean_differences[[i-6]]/as.numeric(MSE[[i]])
  
  # p-value of two-tailed test, rounded to 4 digits
  p_val[[i]] <- round(2*pt(t_ratio[[i]], 8, lower.tail = FALSE), 4)
  
  # Add comparisons to p-values
  p_val[[i]] <- data.frame(comparisons, p_val[[i]])
  
  # rename p value column to p
  colnames(p_val[[i]])[3] = 'p'
  
  # Set values rounded to below 0.0001 to <0.0001
  p_val[[i]]$p[p_val[[i]]$p<0.0001] <- '<0.0001'
  
  # Add columns to set initial polarization ID
  p_val[[i]]$Match <- 'NA'
  
  # Check groups for matching within initial polarization state
  # or within final treatment condition and mark as a match for subsetting
  # For example, comparison of all M0 groups and comparison of M0_Vehicle 
  # with M2_Vehicle is preserved, while M0_Vehicle vs M2_DEP is not preserved
  p_val[[i]] <- p_val[[i]] %>% mutate(Match =
                                        case_when(str_detect(group1, 'M0') &
                                                    str_detect(group2, 'M0') ~ 'Match',
                                                  
                                                  str_detect(group1, 'M2') &
                                                    str_detect(group2, 'M2') ~ 'Match',
                                                  
                                                  str_detect(group1, 'Vehicle') &
                                                    str_detect(group2, 'Vehicle') ~ 'Match',
                                                  
                                                  str_detect(group1, '[\\+]') &
                                                    str_detect(group2, '[\\+]') ~ 'Match',
                                                  
                                                  str_detect(group1, 'M2 -> M1') &
                                                    str_detect(group2, 'M0 -> M1') & 
                                                    !str_detect(group1, '[\\+]') ~ 'Match',
                                                  
                                                  str_detect(group1, '_DEP') &
                                                    str_detect(group2, '_DEP') ~ 'Match',
                                                  
                                                  TRUE ~ as.character(.$Match)))
  
  p_val[[i]] <- subset(p_val[[i]], Match == 'Match')
  
}

# Creating a function to calculate the SEM
stderror <- function(x) sd(x)/sqrt(length(x)) 

# Removing unneeded information from all data, retaining only Treatment ID
Mean <- Fluidigm[-c(1:3, 5:6)] 

# Calculating the average rounded to 2 significant figures +/- SEM
Mean = Mean %>% 
  group_by(Treatment) %>% 
  summarize(across(everything(), 
                   function(x) {
                     paste(round(mean(x),2), 
                           "\u00b1", 
                           round(stderror(x), 2))}))

Mean <- data.frame(Mean[,], row.names = Mean$Treatment) # Making results into a data frame
Mean <- Mean[-c(1)] # Removing treatment column as row names are treatment IDs
names(Mean) <- gsub("\\.", "", names(Mean)) # Removing . from column names
Mean <- t(Mean)
Mean <- as.data.frame(Mean[, c("M0_Vehicle", 'M2_Vehicle', 'M0_DEP', 'M2_DEP',
                               "M0 -> M1", "M2 -> M1", "M0 -> M1+DEP", "M2 -> M1+DEP")])

# Exporting results into an Excel file
write.xlsx(Mean, 'Fluidigm_Mean_SEM.xlsx',
           sheetName = 'Means',
           row.names = TRUE,
           append = FALSE)
```

### Graph log2 fold change data and add statistics calculated above. Each saved graph corresponds to a single gene.

```{r}
# Order for groups on graph
level_order <- c("M0_Vehicle", "M0 -> M1", "M0_DEP", "M0 -> M1+DEP",
                 "M2_Vehicle", "M2 -> M1", "M2_DEP", "M2 -> M1+DEP")

# Set vector for plots to be assigned during loop
myplot <- vector('list', ncol(Fluidigm)) 
y_pos <- vector('list', ncol(Fluidigm))
y_min <- vector('list', ncol(Fluidigm))
y_max <- vector('list', ncol(Fluidigm))
anov_pos <- vector('list', ncol(Fluidigm))

# Identify maximum value of each analyte to set y axis
max <- sapply(Fluidigm, max)
max <- as.numeric(max)

min <- sapply(Fluidigm, min)
min <- as.numeric(min)

# Make a plot for each target in a loop
for(i in 7:ncol(Fluidigm)) {
  
  # If/else to determine y-axis limits and stats y position
  # If the absolute value of the lower limit is greater than the upper limit, 
  # make the lower limit larger and have the stats appear below the data
  # else make the upper limit larger and have stats appear on the top of the data
  if(abs(min[[i]]) > abs(max[[i]])){
    
    anov_pos[[i]] <- as.numeric(min[[i]]*2.75)

    y_min[[i]] <- as.numeric(min[[i]]*2.8)
    y_max[[i]] <- as.numeric(max[[i]]*1.25)
    
    y_pos[[i]] <- c(min[[i]]*2.2, 
                    
                    min[[i]]*1.15, min[[i]]*1.3, min[[i]]*1.45, 
                    min[[i]]*1.15, min[[i]]*1.3, min[[i]]*1.45,
                    
                    min[[i]]*2.35, 
                    
                    min[[i]]*1.6, min[[i]]*1.75, 
                    min[[i]]*1.6, min[[i]]*1.75,
                    
                    min[[i]]*2.5, 
                    
                    min[[i]]*1.9, 
                    min[[i]]*1.9, 
                    
                    min[[i]]*2.65)
                    
  } 
  
  else {
    
    anov_pos[[i]] <- as.numeric(max[[i]]*2.75)

    y_min[[i]] <- as.numeric(min[[i]]*1.25)
    y_max[[i]] <- as.numeric(max[[i]]*2.8)
    
    y_pos[[i]] <- c(max[[i]]*2.2, 
                    
                    max[[i]]*1.15, max[[i]]*1.3, max[[i]]*1.45, 
                    max[[i]]*1.15, max[[i]]*1.3, max[[i]]*1.45,
                    
                    max[[i]]*2.35, 
                    
                    max[[i]]*1.6, max[[i]]*1.75, 
                    max[[i]]*1.6, max[[i]]*1.75,
                    
                    max[[i]]*2.5, 
                    
                    max[[i]]*1.9, 
                    max[[i]]*1.9, 
                    
                    max[[i]]*2.65)
    
  }
  
  # Make one plot per gene, saving it in myplot location [[i]]
  myplot[[i]]<- ggplot(Fluidigm, 
                       aes_string(x = "factor(Treatment, level = level_order)", 
                                  y = Fluidigm[, i])) + 
    
    # Add bars with line at mean
    geom_bar(aes(col=Treatment), 
             position = position_dodge(),
             fill = 'white', 
             stat = "summary", 
             fun = "mean", 
             show.legend = FALSE) +
    
    # Add individual data points
    geom_jitter(aes(col=Treatment), 
                width = 0.25, 
                show.legend = FALSE) + 
    
    # Add a dashed horizontal line at y=0
    geom_hline(yintercept=0, 
               linetype = "dashed") +
    
    # Add anova p-value
    annotate("text", 
             x = 1.5, 
             y = anov_pos[[i]], 
             label = paste0("ANOVA, p = ", anova[[i]])) +
    
    # Add the stats saved in statistics data frame
    stat_pvalue_manual(p_val[[i]], # location where p values are stored
                       y.position = y_pos[[i]], # where on y axis p values/lines should locate
                       label = "p", # name of column in t_test[[i]] where p values stored
                       size = 3.5, # size of labels
                       bracket.size = 0.01, # thickness of lines
                       tip.length = 0, # remove tips from lines
                       vjust = 0.25 # puts p value on comparison line
                       ) +
    
    # Add SEM bars
    stat_summary(geom = "errorbar", 
                 fun.data = mean_se, 
                 position = "dodge", 
                 width = 0.25) +
    
    # y axis limits, calculated in if/else statement
    ylim(y_min[[i]], y_max[[i]]) +
    
    # Label x and y axis
    labs(title = names(Fluidigm)[i],
         x = "Treatment", 
         y = "Log2 Fold Change mRNA") + 
    
    # Set axis label sizing and presentation angle	
    theme(axis.text.x = element_text(face = "bold", size = 14)) + 
    theme(axis.text.x = element_text(angle = 90)) +	
    theme(axis.text.y = element_text(size = 14)) +
    theme(axis.title.y = element_text(size = 16)) +
    theme(axis.title.x = element_text(size = 16)) + 
    theme(panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    
    # Sets legend order to match order of x-axis already on graph due to level_order
    scale_color_discrete(breaks = level_order)
  
  file_name = paste(colnames(Fluidigm)[i], ".tiff", sep="")
  tiff(file_name, units="in", width=8, height=8, res=300)
  print(myplot[[i]])
  dev.off()
  
}
```