---
title: "Improved Graphing Method"
author: "Timothy Smyth"
date: "2024-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Improved Graphing Methods

### This script describes an improved graphing method compared to the method implemented in:

Smyth, Timothy, and Ilona Jaspers. "Diesel exhaust particles induce polarization state-dependent functional and transcriptional changes in human monocyte-derived macrophages." American Journal of Physiology-Lung Cellular and Molecular Physiology 326.1 (2024): L83-L97.

Instead of adding statistics bars and values above or below bars causing the axis to stretch, bars and values are plotted in a separate plot and both plots are combined using patchwork::wrap_plots.

Here, the Fluidigm data is used as an example.

Start by setting up the environment

```{r message = FALSE}
rm(list = ls(all.names = TRUE)) # clears global environ.

# Load packages
library(tidyverse) # for data cleaning
library(pheatmap) # for making heatmap
library(ggfortify) # for PCA
library(factoextra) # for PCA 
library(ggplot2) # for plotting
library(RColorBrewer) # for color palette
library(ggpubr) # for statistics and graphing
library(broom) # for ANOVA
library(ggtext)
library(patchwork)
```

### Load data and format for analysis

```{r}
# Import data
Fluidigm <- read.csv("Fluidigm_Data_for_R.csv", fileEncoding="UTF-8-BOM")
Fluidigm_df <- data.frame(Fluidigm, row.names = Fluidigm$Full.Name)
Fluidigm_df <- Fluidigm_df[, 3:ncol(Fluidigm_df)]

# Isolate Vehicle samples
Veh_df <- subset(Fluidigm, Group == 'Vehicle') 
Veh_df <- data.frame(Veh_df, row.names = Veh_df$Full.Name)
Veh_df <- Veh_df[, 3:ncol(Veh_df)]

# Isolate DEP samples
DEP_df <- subset(Fluidigm, Group == 'DEP') 
DEP_df <- data.frame(DEP_df, row.names = DEP_df$Full.Name)
DEP_df <- DEP_df[, 3:ncol(DEP_df)]

# Remove periods from column names
names(Fluidigm) <- gsub("\\.", "", names(Fluidigm))
names(Fluidigm_df) <- gsub("\\.", "", names(Fluidigm_df))
names(Veh_df) <- gsub("\\.", "", names(Veh_df))
names(DEP_df) <- gsub("\\.", "", names(DEP_df))
```

### Perform 2-way RM ANOVA with fisher's LSD

```{r}
# Create data frame with mean values of each target for testing of differences
mean_df <- Fluidigm_df %>% group_by(Treatment) %>% summarize(across(everything(),mean))
mean_df <- data.frame(mean_df[5:ncol(mean_df)], row.names = mean_df$Treatment)

# Arrange rows as listed
mean_df <- mean_df %>% arrange(factor(rownames(mean_df), 
                                      c("M0_Vehicle", "M2_Vehicle", 
                                        "M0 -> M1", "M2 -> M1",
                                        "M0_DEP", "M2_DEP", 
                                        "M0 -> M1+DEP", "M2 -> M1+DEP")))

# Create row combinations to find differences
Treatment1 <- seq(1,nrow(mean_df)) 
Treatment2 <- Treatment1

# Calculate differences in groups
mean_differences <- expand.grid(Treatment1, Treatment2) %>% # Expand to match sequences
  filter(Var1 > Var2) %>% # Keep rows where Var1 > Var2
  mutate(abs(mean_df[Var1, ] - mean_df[Var2, ])) # Calculate abs value of Var1 - Var2

# Replace var with Treatment name
mean_differences <- mean_differences %>% mutate(Var2 = 
                                                  case_when(str_detect(Var2, '1') ~ "M0_Vehicle",
                                                            str_detect(Var2, '2') ~ "M2_Vehicle",
                                                            str_detect(Var2, '3') ~ "M0 -> M1",
                                                            str_detect(Var2, '4') ~ "M2 -> M1",
                                                            str_detect(Var2, '5') ~ "M0_DEP", 
                                                            str_detect(Var2, '6') ~ "M2_DEP", 
                                                            str_detect(Var2, '7') ~ "M0 -> M1+DEP", 
                                                            str_detect(Var2, '8') ~ "M2 -> M1+DEP"))

mean_differences <- mean_differences %>% mutate(Var1 = 
                                                  case_when(str_detect(Var1, '1') ~ "M0_Vehicle",
                                                            str_detect(Var1, '2') ~ "M2_Vehicle",
                                                            str_detect(Var1, '3') ~ "M0 -> M1",
                                                            str_detect(Var1, '4') ~ "M2 -> M1",
                                                            str_detect(Var1, '5') ~ "M0_DEP", 
                                                            str_detect(Var1, '6') ~ "M2_DEP", 
                                                            str_detect(Var1, '7') ~ "M0 -> M1+DEP", 
                                                            str_detect(Var1, '8') ~ "M2 -> M1+DEP"))

# Isolate comparisons
comparisons <- mean_differences[1:2]
colnames(comparisons) <- c('group1', 'group2')

# Set row names to Treatments compared in row
mean_differences <- data.frame(mean_differences[3:ncol(mean_differences)], 
                               row.names = paste0(mean_differences$Var2, " : ", mean_differences$Var1))

# Create vectors for storing results of stats
Data_Test <- vector('list', ncol(Fluidigm))
anova <- vector('list', ncol(Fluidigm))
MSE <- vector('list', ncol(Fluidigm))
t_ratio <- vector('list', ncol(Fluidigm))
p_val <- vector('list', ncol(Fluidigm))

# Disable scientific notation
options(scipen=999)

# Run Two-Way RM ANOVA with Fisher LSD
# t-ratio of comparison = 

# [Difference between Treatments]/
# sqrt([Mean squares of within group from anova]*(1/n[group1] + 1/n[group2]))

# Individual p-values calculated from the resulting t-ratio with a DF of 8
for(i in 7:ncol(Fluidigm)) {  
  
  # Two-Way RM Anova
  Data_Test[[i]] <- tidy(aov(Fluidigm[, i] ~ Starting * Exposure 
                             + Error(ID / (Starting * Exposure))
                             , data = Fluidigm))
  
  # Isolate anova p-value for graphing
  anova[[i]] <- round(Data_Test[[i]][6, "p.value"], 4)
  
  # Rename column to p
  colnames(anova[[i]])[1] <- 'p'
  
  # Set values rounded to below 0.0001 to <0.0001
  anova[[i]]$p[anova[[i]]$p<0.0001] <- '<0.0001'
  
  # Mean Square Error
  MSE[[i]] <- sqrt(((1/5 + 1/5) * Data_Test[[i]][7,"meansq"]))
  
  # t ratio
  t_ratio[[i]] <- mean_differences[[i-6]]/as.numeric(MSE[[i]])
  
  # p-value of two-tailed test, rounded to 4 digits
  p_val[[i]] <- round(2*pt(t_ratio[[i]], 8, lower.tail = FALSE), 4)
  
  # Add comparisons to p-values
  p_val[[i]] <- data.frame(comparisons, p_val[[i]])
  
  # rename p value column to p
  colnames(p_val[[i]])[3] = 'p'
  
  # Set values rounded to below 0.0001 to <0.0001
  p_val[[i]]$p[p_val[[i]]$p<0.0001] <- '<0.0001'
  
  # Add columns to set initial polarization ID
  p_val[[i]]$Match <- 'NA'
  
  # Check groups for matching within initial polarization state
  # or within final treatment condition and mark as a match for subsetting
  # For example, comparison of all M0 groups and comparison of M0_Vehicle 
  # with M2_Vehicle is preserved, while M0_Vehicle vs M2_DEP is not preserved
  p_val[[i]] <- p_val[[i]] %>% mutate(Match =
                                        case_when(str_detect(group1, 'M0') &
                                                    str_detect(group2, 'M0') ~ 'Match',
                                                  
                                                  str_detect(group1, 'M2') &
                                                    str_detect(group2, 'M2') ~ 'Match',
                                                  
                                                  str_detect(group1, 'Vehicle') &
                                                    str_detect(group2, 'Vehicle') ~ 'Match',
                                                  
                                                  str_detect(group1, '[\\+]') &
                                                    str_detect(group2, '[\\+]') ~ 'Match',
                                                  
                                                  str_detect(group1, 'M2 -> M1') &
                                                    str_detect(group2, 'M0 -> M1') & 
                                                    !str_detect(group1, '[\\+]') ~ 'Match',
                                                  
                                                  str_detect(group1, '_DEP') &
                                                    str_detect(group2, '_DEP') ~ 'Match',
                                                  
                                                  TRUE ~ as.character(.$Match)))
  
  p_val[[i]] <- subset(p_val[[i]], Match == 'Match')
  
}
```

### Graph the data without statistics

```{r}
# Order for groups on graph
level_order <- c("M0_Vehicle", "M0 -> M1", "M0_DEP", "M0 -> M1+DEP",
                 "M2_Vehicle", "M2 -> M1", "M2_DEP", "M2 -> M1+DEP")

# Set vector for plots to be assigned during loop
myplot <- vector('list', ncol(Fluidigm)) 

# Make a plot for each target in a loop
for(i in 7:ncol(Fluidigm)) {
  
  # Make one plot per gene, saving it in myplot location [[i]]
  myplot[[i]]<- ggplot(Fluidigm, 
                       aes(x = factor(Treatment, 
                                      level = level_order), 
                                  y = Fluidigm[, i])) + 
    
    # Add bars with line at mean
    geom_bar(aes(col=Treatment), 
             position = position_dodge(),
             fill = 'white', 
             stat = "summary", 
             fun = "mean", 
             show.legend = FALSE) +
    
    # Add individual data points
    geom_jitter(aes(col=Treatment), 
                width = 0.25, 
                size = 6,
                show.legend = FALSE) + 
    
    # Add a dashed horizontal line at y=0
    geom_hline(yintercept=0, 
               linetype = "dashed") +
    
    # Add SEM bars
    stat_summary(geom = "errorbar", 
                 fun.data = mean_se, 
                 position = "dodge", 
                 width = 0.25) +
    
    # Label x and y axis
    labs(title = NULL,
         x = "Treatment", 
         y = paste0("Log2 Fold Change\n", 
                    colnames(Fluidigm)[i], 
                    ' mRNA')) + 
    
    # Set axis label sizing and presentation angle	
    theme(title = element_text(face = "bold", size = 18),
          axis.text.x = element_text(face = "bold", size = 18, angle = 0),
          axis.text.y = element_text(size = 16),
          axis.title.y = element_text(size = 18),
          axis.title.x = element_text(size = 18),
          panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    
    # Sets legend order to match order of x-axis already on graph due to level_order
    scale_color_discrete(breaks = level_order)
  
}
```

### Create the statistics graphs

The p_val data frames are reorganized by row to match desired layouts. Here, M0 groups, then M2 groups, then M0 vs M2 groups are ordered. 

Then, decide the x axis location for each p-value to be placed. This aims to get the value in the center of the corresponding statistic bar.

Then, decide the y axis location for each p-value to be placed. This aims to place the value right above the statistic bar.

```{r}
anov_pos <- vector('list', ncol(Fluidigm))
mystats <- vector('list', ncol(Fluidigm))

for(i in 7:ncol(Fluidigm)) {
  
  p_val[[i]] <- p_val[[i]][c(2, 3, 4, 9, 10, 14,
                             5, 6, 7, 11, 12, 15,
                             1, 8, 13, 16), ]

  p_val[[i]]$pos_x <- c(1.5, 2, 2.5, 2.5, 3, 3.5,
                        5.5, 6, 6.5, 6.5, 7, 7.5,
                        3, 4, 5, 6)

  p_val[[i]]$pos_y <- seq(from = 0.25,
                          to = 0.25 * nrow(p_val[[i]]),
                          0.25)
  
  # Make one plot per gene, saving it in myplot location [[i]]
  mystats[[i]]<- ggplot(Fluidigm, 
                       aes(x = factor(group2, 
                                      level = level_order), 
                                  y = Fluidigm[, i])) + 
    
    geom_segment(data = p_val[[i]], 
                 aes(x = group1, 
                     xend = group2,
                     y = pos_y, 
                     yend = pos_y), 
                 colour = "black", 
                 linewidth = 1.25) +
    
    geom_text(data = p_val[[i]],
              label = p_val[[i]]$p,
              x = p_val[[i]]$pos_x,
              y = p_val[[i]]$pos_y + 0.1,
              colour = "black") +
    
    # Add anova p-value
    annotate("text",
             size = 8,
             x = 1.5,
             y = max(p_val[[i]]$pos_y),
             label = paste0("ANOVA, p = ", anova[[i]])) +
    
    # Label x and y axis
    labs(title = NULL,
         x = "Treatment", 
         y = paste0(colnames(Fluidigm)[i], "\nLog2 Fold Change mRNA")) + 
    
    # Set axis label sizing and presentation angle	
    theme(title = element_text(face = "bold", size = 18),
          axis.text.x = element_text(face = "bold", size = 18, angle = 0),
          axis.text.y = element_text(size = 16),
          axis.title.y = element_text(size = 18),
          axis.title.x = element_text(size = 18),
          panel.border = element_rect(colour = "black", fill=NA, size=1),
          plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    
    scale_x_discrete(limits = level_order) +
    scale_y_continuous(limits = c(0, max(p_val[[i]]$pos_y) + 0.25))
}
```

### Wrap the data and statistics plots together.

The statistics plot is given theme_void() to remove axis labels and axis to make it not look like one continuous plot.

The data plot has the legend removed. This prevents the statistics bar from extending beyond the data plot which is squished by the legend.

```{r}
for(i in 7:ncol(Fluidigm)){

  tiff(paste0(colnames(Fluidigm)[i], '.tiff'), height = 1000, width = 1500)
  
  print(
  wrap_plots(mystats[[i]] +  
               theme_void(),
             myplot[[i]] + 
               theme(legend.position = "none"), 
             ncol = 1,
             heights = c(1.5, 2)) %>% 
    ggdraw()
  )
  dev.off()
  
}
```